Setting parameters for biological models with ANIMO

### 2014-01-20 Paper.tex
Started to describe parameter support in ANIMO

% - computational modeling in biology
% - assumption of robustness
% - ANIMO, piccolo esempio. ANIMO focuses on topology.
\section{Introduction}
The investigation of biological processes relies on computational support on a daily basis.
This happens not only because of the extremely large amount of data generated in the \emph{``-omics'' era},
but also because many processes are simply too complex to be understood by the human brain alone.
For this reason, systems biology has become more and more important in the last several years.

@@

Proteins are normally expressed at different concentrations in different individuals of the same species, and yet the
overall behavior of their biological networks does not differ significantly. This phenomenon has led to the notion that biological
networks are inherently \emph{robust}. In modeling terms, this means that most of the parameters of a model
can vary inside a certain interval without influencing the qualitative behavior of the whole network.
The interactive approach of ANIMO is based on the assumption of robustness, as our tool is mainly aimed at the
development of network models with a focus on the topology.
Ideally, the biologist can ``play'' with the topology of a network, working
more towards matching the qualitative behavior of experimental data, rather than precisely reproducing it.
However, it is not our intention to concentrate exclusively on the network topology:
in many cases a better parameter choice can improve the behavior of a network more than the addition of new components.
Indeed, making an unnecessarily complex model could reduce its usefulness both in terms of analysis
performances and closeness to reality. The first problem is simply due to the complexity of a network,
which would require more and more computational resources to be analysed\footnote{We also refer to the
problem of \emph{state space explosion}: when a model contains too many loosely coupled components,
the set of its possible evolutions grows exponentially, to the point of making it impossible to apply some analysis
techniques.}.
Realism of network models is more related to their
ultimate usefulness: a model that explains a particular behavior very well but contains many nodes marked
as ``unknown'' has little applicability, as its connection with known processes is very loose.
Therefore, it desirable that a better parameter set is regularly sought for during the design cycle of
a complex biological network model. Some support for parameter choice was already provided in the first versions of ANIMO.
We present here an extended set of tools aimed at achieving a closer fit between ANIMO models and experimental data.
A guideline on how to use these tools to get the best results will also be presented as an ideal workflow.
Thanks to the better awareness on parameter choice gained through this new extension of ANIMO, the biologists
will be able to judge more easily which are the most promising topologies for a network, and thus drive
the experimental research more efficiently.

% - topology - parameters
% - problemi: small example network with parameter variations (also: some parts of a network are more sensitive to parameter variations than others)
\section{ANIMO models}
The starting point of ANIMO is the traditional static representation of biological networks,
which can be easily drawn and managed in softwares like Cytoscape. Indeed, ANIMO was implemented
as a plug-in to Cytoscape, with the aim of adding dynamics to the static representation of biological
networks, and thus allow for analysis on the behavior of such networks. The user interface of ANIMO
can be seen in Figure~\ref{fig:animo-ui}: at the center is the Network panel, where the network
model is represented in the familiar nodes-edges form used in the domain of biology.
Models in ANIMO are \emph{activity-based}, in the sense that nodes have their activity as main
property, and interactions among nodes change the activity of their targets. The concept of activity
is to be intended for example as a generic post-translational modification a molecule can undergo to change its
function. In the case of a kinase, the phosphorylated state is usually interpreted as active.
However, in the context of a gene network the activity of a node standing for a gene represents its current
transcriptional status. A basic type of analysis that can be performed with ANIMO is the generation of
a simulation run, which is presented to the user in the form of a graph (to the right). The activity
graphs generated by ANIMO show the variation in activity of selected nodes over the course of the simulation run.
In addition to that, a slider placed under each graph allows the user to color the nodes in the Network panel
with colors indicating their activity level at any point during the course of the simulation: the Legend
panel on the left links colors to activity.

Of the two main components of an ANIMO model, only the topology is immediately visible to the user in the
Network panel; the parameters are accessed by double clicking the arcs representing node-node interactions.
The dialog window that is shown for an interaction contains the details of the abstract reaction kinetic
describing the interaction, together with the current value of its parameter $k$ (see Fig.~\ref{fig:animo-edge-ui}).
The unique parameter associated to any interaction in ANIMO is used as a scale factor to make the modelled
reaction occur faster or slower. We also provided the user with pre-set values for
$k$, encouraging an initial qualitative assignment of reaction rates as ``slow'', ``fast'' and so forth.
This approach is based on the assumption that, being biological networks inherently robust, the role of
parameters should be perceived as somewhat secondary with respect to the definition of the interactions
in the network. However, as is explained in~\cite{animo-gene}, we still recommend to pay attention to parameter
settings, especially for those parts of a network that tend to be more sensitive to parameter variations.
As can be seen in Figure~\ref{fig:abcd-feedback}, in a feed-back loop the ratio between the involved
parameters plays a determinant role with respect to the behavior of the subnetwork.
%Show example on parameters of feedback inactivation loop (also the example ABCDE from ANIMO's manual would be ok)

% - slow/medium/fast..
% - parameter sweep
% - compare simulations
% Anche come sono fatti (esempio, parameter sweep: la formula dell'errore etc)
\section{Support for parameter synthesis in ANIMO}

>>>>

% - computational modeling in biology
% - assumption of robustness
% - ANIMO, piccolo esempio. ANIMO focuses on topology.
\section{Introduction}
The investigation of biological processes relies on computational support on a daily basis.
This happens not only because of the extremely large amount of data generated in the \emph{-omics era},
but also because many processes are simply too complex to be understood by the human brain alone.
For this reason, systems biology has become more and more important in the last several years.

@@

Proteins are normally expressed at different concentrations in different individuals of the same species, and yet the
overall behavior of their biological networks does not differ significantly. This phenomenon has led to the notion that biological
networks are inherently \emph{robust}. In modeling terms, this means that most of the parameters of a model
can vary inside a certain interval without influencing the qualitative behavior of the whole network.
The interactive approach of ANIMO is based on the assumption of robustness, as our tool is mainly aimed at the
development of network models with a focus on the topology.
Ideally, the biologist can ``play'' with the topology of a network, working
more towards matching the qualitative behavior of experimental data, rather than precisely reproducing it.
However, it is not our intention to concentrate exclusively on the network topology:
in many cases a better parameter choice can improve the behavior of a network more than the addition of new components.
Indeed, making an unnecessarily complex model could reduce its usefulness both in terms of analysis
performances and closeness to reality. The first problem is simply due to the complexity of a network,
which would require more and more computational resources to be analysed\footnote{We also refer to the
problem of \emph{state space explosion}: when a model contains too many loosely coupled components,
the set of its possible evolutions grows exponentially, to the point of making it impossible to apply some analysis
techniques.}.
Realism of network models is more related to their
ultimate usefulness: a model that explains a particular behavior very well but contains many nodes marked
as ``unknown'' has little applicability, as its connection with known processes is very loose.
Therefore, it is desirable that a better parameter set is regularly sought for during the design cycle of
a complex biological network model. Some support for parameter choice was already provided in the first versions of ANIMO.
We present here an extended set of tools aimed at achieving a closer fit between ANIMO models and experimental data.
A guideline on how to use these tools to get the best results will also be presented as an ideal workflow.
Thanks to the better awareness on parameter choice gained through this new extension of ANIMO, the biologists
will be able to judge more easily which are the most promising topologies for a network, and thus drive
the experimental research more efficiently.

% - topology - parameters
% - problemi: small example network with parameter variations (also: some parts of a network are more sensitive to parameter variations than others)
\section{ANIMO models}
The starting point of ANIMO is the traditional static representation of biological networks,
which can be easily drawn and managed in softwares like Cytoscape. Indeed, ANIMO was implemented
as a plug-in to Cytoscape, with the aim of adding dynamics to the static representation of biological
networks, and thus allow for analysis on the behavior of such networks. The user interface of ANIMO
can be seen in Figure~\ref{fig:animo-ui}: at the center is the Network panel, where the network
model is represented in the familiar nodes-edges form used in the domain of biology.
Models in ANIMO are \emph{activity-based}, in the sense that nodes have their activity as main
property, and interactions among nodes change the activity of their targets. The concept of activity
can be intended for example as a generic post-translational modification a molecule can undergo to change its
function. In the case of a kinase, the phosphorylated state is usually interpreted as active.
In the context of a gene network, the activity of a node standing for a gene represents its current
transcriptional status. A basic type of analysis that can be performed with ANIMO is the generation of
a simulation run, which is presented to the user in the form of a graph (to the right). The activity
graphs generated by ANIMO show the variation in activity of selected nodes over the course of the simulation run.
In addition to that, a slider placed under each graph allows the user to color the nodes in the Network panel
depending on their activity level at any point during the course of the simulation: the Legend
panel on the left links colors to activity.

Of the two main components of an ANIMO model, only the topology is immediately visible to the user in the
Network panel; the parameters are accessed by double clicking the arcs representing node-node interactions.
The dialog window that is shown for an interaction contains the details of the abstract reaction kinetic
describing the interaction, together with the current value of its parameter $k$ (see Fig.~\ref{fig:animo-edge-ui}).
The unique parameter associated to any interaction in ANIMO is used as a scale factor to make the modelled
reaction occur faster or slower. We also provide the user with pre-set values for
$k$, encouraging an initial qualitative assignment of reaction rates as ``slow'', ``fast'' and so forth.
This approach is based on the assumption that biological networks are inherently robust:
once an acceptable set of parameters is found, a more precise parameter search will generally have
little impact on the fit of a model.
However, as can be seen in the small example in Figure~\ref{fig:small-example-feedback},
robustness does not imply that any parameter choice will do. In particular, network motifs
such as feedback loops make models more dependent on parameter settings.
As biological networks tend to heavily rely on cross-talk, network topologies can rapidly become
complex. Some manual parameter fitting is currently necessary for the more complex ANIMO models.
This work can be slow and error-prone, taking away some of the user-friendliness
for which ANIMO aims.
We will show how ANIMO has been improved to make parameter choice easier for the user.

\begin{figure}
%Show example on parameters of feedback inactivation loop (also the example ABCDE from ANIMO's manual would be ok)
\caption{fig:small-example-feedback}
\end{figure}

% - slow/medium/fast..
% - parameter sweep
% - compare simulations
% Anche come sono fatti (esempio, parameter sweep: la formula dell'errore etc)
\section{Support for parameter synthesis in ANIMO}
When a model does not match experimental data, ANIMO offers three main tools to achieve a better fit:
manual parameter editing, comparisons between different model versions, and the newly introduced parameter sweep.

\subsection{Manual parameter editing}\label{sec:parameter-editing}
Double-clicking on an arrow in the Cytoscape representation of an interaction lets the user access
a dialog that allows to change the approximated scenario and parameters for the selected interaction.
The simplified scenarios were described in~\cite{animo-ieee}, and we explain them briefly here:
\begin{description}
  \item scenario 1
  \item scen 2
  \item scen 3
\end{description}

The choice for the parameter $k$ can be made directly inserting a numeric value, or by choosing
a preset among the proposed qualitative values \emph{very slow, slow, medium, fast, very fast}.
We advise our users to base their initial parameter choice on these qualitative presets.
In Figure~\ref{fig:small-example-feedback}, all parameters are initially set to \emph{medium}
(Fig.~\ref{fig:small-example-feedback}{\sf a}), while the peak in {\sf b} was obtained by setting the
parameter of $C \dashv B$ to \emph{fast} and the parameter of $E \dashv D$ to \emph{slow}.

\subsection{Comparing model versions}\label{sec:comparing-versions}
Particularly large networks make it more difficult for the user to understand the effect
of a change in network topology or interaction parameters. To overcome this difficulty, ANIMO
allows the user to visually compare two versions of a model in terms of their simulation results.
The Results panel in Figure~\ref{animo-ui} shows the graph of selected node activities during
the course of an ANIMO simulation. The button ``Compare with\dots'' allows to compare the current
simulation data with another based on a possibly different version of the model. When the two
simulations to be compared have been chosen, ANIMO produces a new graph plotting the difference
between the two original simulations. The slider under the new graph allows to visualize also
in the Network panel the changes in node activity in the whole network: Figure~\ref{fig:diff}
shows the difference between the two versions of the model in Figure~\ref{fig:small-example-feedback}.
The difference was computed as {\sf b - a}. Nodes colored in green in the Network panel are
more active in version {\sf b}, while red nodes are more active in version {\sf a}.

By changing the title of a simulation, a user can keep track of different versions of their model,
which can be also kept and saved with the model, to be used in future sessions or for sharing purposes.
Should a user want to backtrack to a particular version of the network, the ``Reset to this'' button
can be used. As an additional help to recall what a model version consisted of, the tooltipo of
the ``Reset to this'' button shows an image of the network topology, taken in the moment when
the selected simulation was begun. A click on the ``Reset to this'' button will reset the
network topology and parameters at the ones used to generate the selected simulation.
As the adaptation of a model may proceed on different paths, the ``Compare with\dots'' and
``Reset to this'' buttons provide the user with some help in tracking the changes to the model
and selecting the most promising ones.

\subsection{Parameter sweep}\label{sec:parameter-sweep}
In order to avoid introducing too many changes to the topology of a network, we recommend
ANIMO users to be reasonably sure that a topology does not fit a given data set before
modifying it. As this would entail an extensive parameter search which, even with the tools
described in Sections~\ref{sec:parameter-editing} and~\ref{sec:comparing-versions}, would
require a considerable amount of time and effort, we have provided the latest version of ANIMO
with a support for parameter sweeps. The idea is to let ANIMO explore the user-defined parameter
space by computing a series of simulations based on the same network topology, when only the
parameters are being changed. Comparing the results with a given experimental data series
will let the user select the most promising parameter choices, eventually selecting the one
that behaves the best.
%Multicore!
%log e linear scale
%error vs data --> narrower parameter choices
